CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-22 docs/CrystalPlan_paper/paper.tex
Refs #182: RMS stats for peak offsets, more paper writing.

\documentclass[final]{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage{graphicx}
%\usepackage[pdftex]{graphicx}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\Xi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the $\lambda_{min}$ and $\lambda_{max}$ wavelength limits
of the detectors and/or source. The scattered beam direction and the two possible incident beam vectors
are used to to find two $q$ vectors, which point in the same direction. At the
high wavelength we find $q_{min}$, and the low wavelength limit scattering off
of $q_{max}$. Since the instrument has a wide bandwidth, we fill in the 3D q-space
matrix by filling all points from $q_{min}$ to the $q_{max}$, for all pixels in
the face of the detector. This quickly defines a volume of coverage, as shown
schematically in two dimensions in Figure \ref{fig:ewald_fig}.

@@

@@

@@

\begin{table}
\caption{Caption to table}
\begin{tabular}{llcr}      % Alignment for each cell: l=left, c=center, r=right
 HEADING    & FOR        & EACH       & COLUMN     \\
\hline
 entry      & entry      & entry      & entry      \\
 entry      & entry      & entry      & entry      \\
 entry      & entry      & entry      & entry      \\
\end{tabular}
\end{table}

@@

\begin{figure}
\caption{Ewald sphere construction showing reciprocal-space detector coverage
for a wide-bandwidth Laue diffractometer. The smaller blue circle represents the
$1/\lambda_{max}$ limit of measurement, with the larger green circle
representing $1/\lambda_{min}$. The origin of reciprocal space (000) is common to both
spheres, but the sample position in the construction (S1 or S2) depends on
$\lambda$. The angular coverage of the detector is represented by the dashed
wedge, and projects onto each sphere of radius $1/\lambda$ an arc. The red wedge
between the two $\lambda$ is built by projecting straight lines between the
$q_{min}$ and $q_{max}$ limits, giving the volume in reciprocal space that is
measured at a single sample oriention. }

>>>>

\documentclass[draft]{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage[version=3]{mhchem} 
%\usepackage[pdftex]{graphicx}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\chi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the $\lambda_{min}$ and $\lambda_{max}$ wavelength limits
of the detectors and/or source. The scattered beam direction and the two possible incident beam vectors
are used to to find two $q$ vectors, which point in the same direction. At the
high wavelength we find $q_{min}$, and the low wavelength limit scattering off
of $q_{max}$. Since the instrument has a wide bandwidth, we fill in the 3D q-space
matrix by filling all points from $q_{min}$ to the $q_{max}$, for all pixels in
the face of the detector. This quickly defines a volume of coverage, as shown
schematically in two dimensions in Figure \ref{fig:ewald_fig}.

@@

%---------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------

@@

As part of the TOPAZ beamline commissioning process, a Natrolite
(\ce{Na2Al2Si3O10} $\cdot$ \ce{2H2O}) single crystal was measured on a fixed
$45\degree\ \chi$ goniometer at room temperature and ambient atmosphere. An initial short run was processed using
ISAW to determine the UB matrix and cell parameters of the crystal.

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations that covered 95\% of the peaks, taking into account the crystal
symmetry (orthorombic or mmm). The plan was left to run unattended overnight.

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Very good agreement was found; the
RMS difference between the predicted and actual measurement wavelength was only
0.016 $\AA$, whereas the RMS of the detector position error was
7.6 mm. Inspecting the offsets showed that the main contributor to this error was a
slight misalignment of the goniometer, which CrystalPlan can help diagnose.

%NOTE: Check on this misalignment of goniometer business - is that really the
% cause?

%---------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------

@@

@@

\begin{figure}
\caption{2D Ewald sphere construction showing reciprocal-space detector coverage
for a wide-bandwidth Laue diffractometer. The smaller blue circle represents the
$1/\lambda_{max}$ limit of measurement, with the larger green circle
representing $1/\lambda_{min}$. The origin of reciprocal space (000) is common to both
spheres, but the ``virtual" sample position in the construction (S1 or S2)
depends on $\lambda$. The angular coverage of the detector is represented by the dashed
wedge, and projects an arc onto each sphere of radius $1/\lambda$. The
red wedge between the two $\lambda$ is built by projecting straight lines between the
$q_{min}$ and $q_{max}$ limits for each pixel on the detector face, giving the
volume in reciprocal space that is measured at a single sample oriention. }
